Changing face of pediatric pulmonary exacerbations in CF

### 2023-03-29 Scott Sagel/PEx in CF/nacfc_2023.qmd
Update nacfc_2023.qmd

---
title: "Changing face of pediatric pulmonary exacerbations in CF"
author: "Tim Vigers"
date: "today"
format:
  pdf:
    toc: true
    lot: true
    lof: true
editor: source
execute: 
  echo: false
---

```{r}
#| include: false
library(readxl)
library(tidyverse)
library(gtsummary)
source("~/GitHub/CF/genotype_class_severity.R")
# Import
chco = read_excel("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Scott Sagel/PEx in CF/Data_Clean/CHCO.xlsx",na = c("","NULL"))
cnh = read_excel("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Scott Sagel/PEx in CF/Data_Clean/CNH.xlsx")
# Format factor columns
chco$`Modulator at Time of Admit` = factor(chco$`Modulator at Time of Admit`,
                                           levels = c("No","Yes"),ordered = T)
```

@@

1. Age at admission and days in hospital were averaged for each person with multiple admissions.

```{r}
#| label: tbl-chco
#| tbl-cap: CHCO Summary by Year
#| message: false
chco %>% group_by(`Unique number`,Year) %>%
  summarise(`# of Hospitalizations per Person` = n(),
            across(Sex:Genotypes2,~first(na.omit(.x))),
            across(`Modulator at Time of Admit`,~max(.x)),
            across(all_of(c("Age at Admit","# Days in Admit")),~mean(.x,na.rm=T)),
            .groups = "drop") %>%
  mutate(Genotypes1 = case_when(
    Genotypes1 == "F508del" | Genotypes1 == "delF508" ~ "F508del",
    .default = "Other"
  ),
  Genotypes2 = case_when(
    Genotypes2 == "F508del" | Genotypes2 == "delF508" ~ "F508del",
    .default = "Other"
  )) %>%
  unite(Genotype,Genotypes1,Genotypes2,sep = "/") %>%
  mutate(Genotype = factor(Genotype,
                           levels=c("F508del/F508del","F508del/Other",
                                             "Other/F508del","Other/Other"),
                           labels = c("F508del/F508del","F508del/Other",
                                             "F508del/Other","Other/Other")
                           )) %>%
  select(`# of Hospitalizations per Person`,`# Days in Admit`,`Age at Admit`,
         `Modulator at Time of Admit`,Sex:Genotype,Year) %>%
  tbl_summary(by=Year) %>%
  add_p() %>%
  as_gt()
```

The number of hospitalizations for PEx decreased from `r sum(chco$Year==2018)` in 2018, involving `r nrow(unique(chco[chco$Year==2018,"Unique number"]))` unique individuals, to `r sum(chco$Year==2022)` in 2022, involving `r nrow(unique(chco[chco$Year==2022,"Unique number"]))` unique individuals.

>>>>

---
title: "Changing face of pediatric pulmonary exacerbations in CF"
author: "Tim Vigers"
date: "today"
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
editor: source
execute: 
  echo: false
---

```{r}
#| include: false
library(readxl)
library(tidyverse)
library(gtsummary)
source("~/GitHub/CF/genotype_class_severity.R")
# Import
chco = read_excel("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Scott Sagel/PEx in CF/Data_Clean/CHCO.xlsx",na = c("","NULL"))
cnh = read_excel("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Scott Sagel/PEx in CF/Data_Clean/CNH.xlsx")
# Format CHCO columns
chco$`P. aeruginosa` = coalesce(chco$`P. aeruginosa mucoid`,chco$`P. aeruginosa non-mucoid`)
chco = chco %>% rename(
  Race = `Race (White/Black/Other)`,Ethnicity = `Hispanic/Latinx`
) %>%
  mutate(Genotypes1 = case_when(
    Genotypes1 == "F508del" | Genotypes1 == "delF508" ~ "F508del",
    .default = "Other"
  ),
  Genotypes2 = case_when(
    Genotypes2 == "F508del" | Genotypes2 == "delF508" ~ "F508del",
    .default = "Other"
  ),
  `FEV1 % Predicted at Admit` = suppressWarnings(as.numeric(`FEV1 % Predicted at Admit`))) %>%
  unite(Genotype,Genotypes1,Genotypes2,sep = "/")
# Format CNH to match CHCO
cnh = cnh %>%
  rename(`# Days in Admit`=`Total hospitalization days`,
         `Age at Admit`=`Age at admission`,
         `FEV1 % Predicted at Discharge`=`End of treamtent ppFEV1`,
         `Unique number` = `Study ID`) %>%
  mutate(Site = "CNH",
         `Modulator at Time of Admit` = case_when(
           None == "Unchecked" ~ "Yes",
           None =="Checked" ~"No"
         ),
         MRSA = case_when(
           MRSA == "Unchecked" ~ NA,
           MRSA =="Checked" ~ TRUE
         ),
         `Staph aureus` = case_when(
           MSSA == "Unchecked" ~ NA,
           MSSA =="Checked" ~ TRUE
         ),
         `P. aeruginosa` = case_when(
           `P. aeruginosa` == "Unchecked" ~ NA,
           `P. aeruginosa` =="Checked" ~ TRUE
         ),
         Genotype = case_when(
           `CFTR genotype, groups` =="F508del heterozygous" ~ "F508del/Other",
           `CFTR genotype, groups` =="F508del homozygous"~"F508del/F508del",
           `CFTR genotype, groups` =="Other"~"Other/Other"
         ),
         `Specific Modulator at Admit` = case_when(
           `Ivacaftor (Kalydeco)` == "Checked" ~ "Other",
           `Ivacaftor/lumacaftor (Orkambi)` == "Checked" ~ "Other",
           `Tezacaftor/ivacaftor (Symdeko)`=="Checked" ~ "Other",
           `Elexacaftor/tezacaftor/ivacaftor (Trikafta)`=="Checked" ~ "Trikafta",
           .default = "None"
         ),
         `FEV1 % Predicted at Admit` = as.numeric(`PEx ppFEV1`))
# Merge
df = full_join(chco,cnh)
# Format column types
df$`Modulator at Time of Admit` = factor(df$`Modulator at Time of Admit`,
                                         levels = c("No","Yes"),ordered = T)
df$`Specific Modulator at Admit`[df$`Modulator at Time of Admit`=="No"] = "None"
df$`Specific Modulator at Admit` = factor(tolower(df$`Specific Modulator at Admit`),
                                          levels = c("kalydeco","none","orkambi","symdeko","trikafta"),
                                          labels = c("Other","None","Other","Other","Trikafta"))
df$Race = factor(df$Race,levels = c("Amer. Indian or Alaskan Naitive","Asian",
                                    "Black","Black or African American","Black/ AA",
                                    "Decline to Answer","Hispanic","Other",
                                    "Unknown / Not Reported","White"),
                 labels = c("Other","Other","Black","Black","Black",
                            "Unknown / Not Reported","Other","Other",
                            "Unknown / Not Reported","White"))
df$Ethnicity = factor(df$Ethnicity,levels = c("FALSE","Hispanic or Latino","HLS",
                                              "NOT Hispanic or Latino","NOT HLS",
                                              "TRUE","UNKNOWN"),
                      labels = c("NOT Hispanic or Latino","Hispanic or Latino",
                                 "Hispanic or Latino","NOT Hispanic or Latino",
                                 "NOT Hispanic or Latino","Hispanic or Latino",
                                 "Unknown"))
bugs = c("P. aeruginosa","P. aeruginosa mucoid","P. aeruginosa non-mucoid","Staph aureus","MRSA")
df[,bugs] = lapply(df[,bugs],function(c){
  c[is.na(c)] = "No"
  c = factor(c,levels = c("No",TRUE),labels = c("No","Yes"),ordered = T)
})
# Make summary table
summ = df %>% group_by(`Unique number`,Site,Year) %>%
  mutate(`Change in FEV1` = `FEV1 % Predicted at Discharge` - `FEV1 % Predicted at Admit`) %>%
  summarise(`# of Hospitalizations per Person` = n(),
            across(c(Sex:Genotype,`Specific Modulator at Admit`),~first(na.omit(.x))),
            across(c(`Modulator at Time of Admit`,all_of(bugs)),~max(.x,na.rm = T)),
            across(all_of(c("Age at Admit","# Days in Admit","Change in FEV1",
                            "FEV1 % Predicted at Admit","FEV1 % Predicted at Discharge")),
                   ~mean(.x,na.rm=T)),
            .groups = "drop")  %>%
  mutate(Genotype = factor(Genotype,
                           levels=c("F508del/F508del","F508del/Other",
                                    "Other/F508del","Other/Other"),
                           labels = c("F508del/F508del","F508del/Other",
                                      "F508del/Other","Other/Other")
  )) %>%
  select(`Unique number`, `Site`,Year,`# of Hospitalizations per Person`,`# Days in Admit`,`Age at Admit`,
         `Change in FEV1`,`FEV1 % Predicted at Admit`,`FEV1 % Predicted at Discharge`,
         `Modulator at Time of Admit`,`Specific Modulator at Admit`,
         all_of(bugs),Sex:Genotype)
```

@@

1. For each person with multiple admissions, continuous variables were averaged prior to generating summary tables.
2. If a participant was on a modulators during any admission, they were counted as on modulators for that year. For example, participant 4 from CHOC was hospitalized 3 times in 2018 and was on a modulator during one of those admissions, so they were counted as on modulators in 2018.
3. CNH only included combined P. aeruginosa information.

# Summary Statistics

```{r}
#| label: tbl-chco
#| tbl-cap: Summary by Year
#| message: false
summ %>%
  select(-`Unique number`,-Site) %>%
  tbl_summary(by=Year,
              type = list(`# of Hospitalizations per Person` ~ 'continuous'),
              statistic = list(
                c("# of Hospitalizations per Person","# Days in Admit") ~ "{median} ({p25}, {p75})",
                c("Age at Admit","Change in FEV1","FEV1 % Predicted at Admit",
                  "FEV1 % Predicted at Discharge") ~ "{mean} ({sd})"
              ),
              missing_text = "(Missing)") %>%
  add_ci(pattern = "{stat} [{ci}]") %>%
  add_p(test = list(
    `Age at Admit`~"t.test",
    `Change in FEV1`~"t.test"
  )) %>%
  separate_p_footnotes() %>%
  as_gt()
```

The number of hospitalizations for PEx decreased from `r sum(summ$Year==2018)` in 2018 to `r sum(summ$Year==2022)` in 2022.
